Analysis for Project

### 2021-11-17 analysis/analysis.Rmd
Switch from predict to modmarg to calculate marginal means
The results will change slightly because I am now estimating means averaged across all respondents rather than at the mean respondent.

All models are estimated as logit models. In order to calculate conditional means for the plots below, I need to be able to feed back in data to the `predict` command, including fractional values for 1/0 dummy variables. Therefore, rather than use the `acs` data itself, I need to create a purely numeric design matrix where each factor variable is expanded out into a set of 1/0 dummies.

```{r design-matrix}

metro <- model.matrix(~metro, acs)[,-1]
colnames(metro) <- c("met_non_metro","met_central_city","met_non_central_city",
                     "met_city_indeter")
#mean center
metro <- t(t(metro)-apply(metro, 2, mean))
metro <- as.data.frame(metro)

years <- model.matrix(~as.factor(year), acs)[,-1]
colnames(years) <- paste("y", 2011:2019, sep="")
#mean center
years <- t(t(years)-apply(years, 2, mean))
years <- as.data.frame(years)

states <- model.matrix(~as.factor(state), acs)[,-1]
colnames(states) <- paste("s", 1:50, sep="")
#mean center
states <- t(t(states)-apply(states, 2, mean))
states <- as.data.frame(states)

grades <- model.matrix(~current_grade, acs)[,-1]
#mean center
grades <- t(t(grades)-apply(grades, 2, mean))
grades <- as.data.frame(grades)

degree <- model.matrix(~degree_mother+degree_father, acs)[,-1]
#mean center
degree <- t(t(degree)-apply(degree, 2, mean))
degree <- as.data.frame(degree)

spk_eng <- model.matrix(~spk_eng+spk_eng_father+spk_eng_mother, acs)[,-1]
#mean center
spk_eng <- t(t(spk_eng)-apply(spk_eng, 2, mean))
spk_eng <- as.data.frame(spk_eng)

design_matrix <- cbind(states, years, grades, metro, degree, spk_eng)

design_matrix$foreign_born <- acs$foreign_born-mean(acs$foreign_born)
design_matrix$foreign_born_mother <- acs$foreign_born_mother-
  mean(acs$foreign_born_mother)
design_matrix$foreign_born_father <- acs$foreign_born_father-
  mean(acs$foreign_born_father)

design_matrix$income_trans <- sqrt(acs$family_income/10000)
design_matrix$income_trans <- design_matrix$income_trans-
  mean(design_matrix$income_trans)
design_matrix$own_home <- acs$own_home-mean(acs$own_home)
design_matrix$parents_married <- acs$parents_married-
  mean(acs$parents_married)

design_matrix$race <- acs$race
design_matrix$below_exp_grade <- acs$below_exp_grade

# add design effects
design_matrix$cluster <- acs$cluster
design_matrix$strata <- acs$strata
design_matrix$perwt <- acs$perwt

design_matrix <- svydesign(ids=~cluster, data=design_matrix)
```

Now, I can run the models. I am using the `svyglm` command which (given the design set up above) adjusts for clustering and weights observations based on sample weights. I also want to figure out how to adjust for strata, but `svydesign` is throwing up errors when I attemp to do that. Warning, the code chunks below take a bit of time to run!

```{r build-models, results='asis'}
model0 <- svyglm(below_exp_grade~race, design=design_matrix, family=binomial)
model1 <- update(model0, 
                 .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
                   current_grade1st+current_grade2nd+current_grade3rd+
                   current_grade4th+current_grade5th+current_grade6th+
                   current_grade7th+current_grade8th+current_grade9th+
                   current_grade10th+current_grade11th+current_grade10th+
                   current_grade11th+current_grade12th+
                   met_non_metro+met_central_city+met_non_central_city+
                   met_city_indeter+
                   s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+
                   s18+s19+s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+
                   s33+s34+s35+s36+s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+
                   s48+s49+s50)
model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
                   foreign_born_father+spk_engSomewhat+spk_engYes+
                   spk_eng_motherSomewhat+spk_eng_motherYes+
                   spk_eng_fatherSomewhat+spk_eng_fatherYes)
model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
                   degree_motherG+degree_fatherHS+degree_fatherAA+
                   degree_fatherBA+degree_fatherG)
model4 <- update(model3, .~.+income_trans+own_home+parents_married)

@@

I can do this by calculating conditional means (or proportions to be more accurate) of the dependent variable across all of the possible race groups. For each conditional mean, I can also calculate standard errors for these estimates by the delta method. This can all be done via the `predict` command.

To compare the results for each biracial gropu to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value. Calculating the midway value is just a straightforward mean calculation. To calculate the standard error, I take:

@@

There are two additional issues that have to be addressed though. First, I need to decide whether I want to calculate conditional means at the level of the response (proportions) or the linear transformation (log-odds). For ease of interpretability, I think the best option here is the former for the main presentation of results, but I also want to test the sensitivity of these results if I do it the other way.

Second, the conditional means only apply at specific values of the independent variables. So I need to hold all of the other variables at some value. Now the relative difference between racial groups wouldn't change if this was a linear model (or if I used log-odds rather than proportions) so those values wouldn't matter much in that case. However, if I am estimating the actual proportions, then I no longer have a linear model and I am now caught up indirectly in the *logit panic*. I don't think it will matter that much because the range of predicted values of grade retention is not going to be that high (like between 2% and 10%). However, I should test the sensitivity of the results to different specifications. As a default, it makes sense to use the mean values for all other variables. Thankfully, because I already mean-centered the design matrix, this means that all other variables should be at zero.

```{r create-predictor-data}
#dateset for predict
df_pred <- data.frame(race=levels(acs$race),
                      y2011=0, y2012=0, y2013=0, y2014=0, y2015=0, y2016=0,
                      y2017=0, y2018=0, y2019=0,
                      current_grade1st=0, current_grade2nd=0, current_grade3rd=0,
                      current_grade4th=0, current_grade5th=0, current_grade6th=0,
                      current_grade7th=0, current_grade8th=0, current_grade9th=0,
                      current_grade10th=0, current_grade11th=0, current_grade12th=0,
                      met_non_metro=0, met_central_city=0, met_non_central_city=0,
                      met_city_indeter=0,
                      s1=0, s2=0, s3=0, s4=0, s5=0, s6=0, s7=0, s8=0, s9=0, s10=0,
                      s11=0, s12=0, s13=0, s14=0, s15=0, s16=0, s17=0, s18=0, s19=0,
                      s20=0, s21=0, s22=0, s23=0, s24=0, s25=0, s26=0, s27=0, s28=0,
                      s29=0, s30=0, s31=0, s32=0, s33=0, s34=0, s35=0, s36=0, s37=0,
                      s38=0, s39=0, s40=0, s41=0, s42=0, s43=0, s44=0, s45=0, s46=0,
                      s47=0, s48=0, s49=0, s50=0,
                      degree_motherHS=0, degree_motherAA=0, degree_motherBA=0,
                      degree_motherG=0, degree_fatherHS=0, degree_fatherAA=0, 
                      degree_fatherBA=0, degree_fatherG=0,
                      foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
                      spk_engSomewhat=0, spk_engYes=0, spk_eng_fatherSomewhat=0, 
                      spk_eng_fatherYes=0, spk_eng_motherSomewhat=0, spk_engYes=0,
                      income_trans=0, own_home=0, parents_married=0)
```

Now in order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.

```{r calc-cond-means}
coef_table <- calculate_cond_means(model0)
```

@@

```{r example-plot}
g <- coef_table %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial black/white respondents and monoracial\ncomparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")

g+annotate("text", 1, 0.0375, hjust="left",
             label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.048, hjust="left",
             label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.039, hjust="left",
             label="Actual biracial estimate\nshown with 83% confidence\ninterval")
```

This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors. The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty. I can then see where the actual biracial estimate falls within this range. The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students. However, their probability is still a bit higher than that for whites.

@@

```{r cond-means-model4}
coef_table <- calculate_cond_means(model4)

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

```{r cond-means-model-compare, fig.height=12}
coef_table <- calculate_cond_means(model0) %>%
  mutate(model="crude")
coef_table <- calculate_cond_means(model1) %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model2) %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model4) %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

>>>>

All models are estimated as logit models. Warning, the code chunks below take a bit of time to run!

```{r build-models, results='asis'}
model0 <- glm(below_exp_grade~race, data=acs, family=binomial)
model1 <- update(model0, as.factor(year)+current_grade+as.factor(state)+metro)
model2 <- update(model1, 
                 .~.+foreign_born+foreign_born_mother+foreign_born_father+
                   spk_eng+spk_eng_mother+spk_eng_father)
model3 <- update(model2, .~.+degree_mother+degree_father)
model4 <- update(model3, .~.+sqrt(family_income/10000)+own_home+parents_married)

@@

I can do this by calculating marginal means (or proportions to be more accurate) of the dependent variable across all of the possible race groups. In order to calculate these marginal means, I use the `marg` command in the `modmargs` library. This gives me results equivalent to average marginal effects but on the level (e.g. mean) rather than the effect (different in mean). Standard errors for these marginal means are calculated by the [delta method](https://cran.r-project.org/web/packages/modmarg/vignettes/delta-method.html). I adjust all SEs for clustering of respondents within households.

To compare the results for each biracial group to the halfway assumption, I need to calculate the value midway between the two consituent groups and the standard error of that value. Calculating the midway value is just a straightforward mean calculation. To calculate the standard error, I take:

@@

In order to calculate all the values that I need, I have created a function that will put this into a nice table for later plotting.

```{r calc-cond-means, warning=FALSE}
coef_table_mod0 <- calculate_cond_means(model0)
```

@@

```{r example-plot}
g <- coef_table_mod0 %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_mod0, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial black/white respondents and monoracial\ncomparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")

g+annotate("text", 1, 0.035, hjust="left",
             label="Grey bands give 83.4%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.045, hjust="left",
             label="Red band gives 83.4%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.036, hjust="left",
             label="Actual biracial estimate\nshown with 83.4% confidence\ninterval")
```

This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors (although the SEs are slightly different because this model accounts for clustering). The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty. I can then see where the actual biracial estimate falls within this range. The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students. However, their probability is still a bit higher than that for whites.

@@

```{r cond-means-model4}
coef_table_mod4 <- calculate_cond_means(model4)

ggplot(coef_table_mod4, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table_mod4, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

```{r cond-means-model-compare, fig.height=12}
#calculate remaining marginal means
coef_table_mod1 <- calculate_cond_means(model1)
coef_table_mod2 <- calculate_cond_means(model2)

coef_table <- coef_table_mod0 %>%
  mutate(model="crude")
coef_table <- coef_table_mod1 %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- coef_table_mod2 %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- coef_table_mod4 %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))
